---
title: "MACT6100 Assessment 2"
author: "Junchi Han"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
options(width = 65)
```

# 1. Introduction

## 1.1 Introduction to Customer Churn

Customer churn is a major challenge in the telecommunications industry due to
intense competition and low switching costs. Existing research shows that churn rates
are directly related to the revenue. It is also suggested that retaining customers
is more cost-effective than acquiring new ones. As a result, churn prediction
has become a key application of machine learning and statistical modelling in
both academia and industry.

## 1.2 Background of Customer Churn Data

Previous academic studies has demonstrated that classification methods can effectively
predict customer churn using behavioural and demographic data. This study uses
the Telco Customer Churn dataset from Kaggle, which contains rich information on
customer demographics, service usage, contracts, and billing behaviour.
The topic was chosen due to its strong real-world relevance and its suitability
for comparing machine-learning techniques in a practical setting.

## 1.3 Aim of the Research

This study aims to address the following core questions: how do different models
(GLM, Decision Tree, Neural Network, and Random Forest) compare in their performance
for churn prediction, and which model performs best? In addition, within the Telco
Customer Churn dataset, which features are most important for predicting whether
a customer will churn?

## 1.4 Methology & Models

To address the research questions, this project applies a range of statistical
and machine learning methods to build and compare predictive models. 

The methodology includes data cleaning, exploratory data analysis (EDA), train-test
split, and model performance evaluation.

The models considered include the traditional statistical approach of GLM, as well
as machine learning and deep learning models such as Decision Tree, Neural Network,
and Random Forest.
All models are evaluated and compared using performance metrics including ROC,
AUC and accuracy, to provide a comprehensive assessment of their effectiveness in
customer churn prediction.

# 2. Data Loading & Exploring

## 2.1 Load Customer Churn Data

"Telco Customer Churn" data is downloaded from https://www.kaggle.com/datasets/blastchar/telco-customer-churn?resource=download

```{r}
library(tidyverse)
sdata = read.csv('WA_Fn-UseC_-Telco-Customer-Churn.csv')
str(sdata)
```
It is suggested that there are 7043 observations of  21 variables in the original churn data.
The data are stored in different categories such as character, number or integer.

It is also worthy check if data contains any N/A values.

```{r}
sum(is.na(sdata$TotalCharges))
```

```{r}
sdata = na.omit(sdata)
sum(is.na(sdata$TotalCharges))
```
Noticed that 11 N/A values have been removed.

It is sensible to remove "customerID" column, as they are unique identifier of each
of the customer and are not useful for prediction.
```{r}
sdata <- sdata %>% dplyr::select(-customerID)
```



















